---
title: "Project 3"
format: html
---

**Completed by**: Amanda Kim (akim148) and Da Yea Song (dsong13)

```{r}
suppressPackageStartupMessages(library(caret))
suppressPackageStartupMessages(library(DBI))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(nycflights13))
suppressPackageStartupMessages(library(pROC))
suppressPackageStartupMessages(library(RSQLite))
suppressPackageStartupMessages(library(viridis))
```

## Part 1

In this part, you and your partner will use the `DBI` and `RSQLite` packages to write SQL queries, explore and summarize the data, visualize the data, and summarize your findings.

Using the data you picked, choose a question to investigate. For example, if you are using the `nycflights13` dataset, maybe you want to understand if there are certain days of the week that flights are more delayed than other days of the week. Describe what is the question you aim to answer with the data and what you want to visualize.

1.  **If it's already a SQL database, read into R using the `DBI` and `RSQLite` packages. If it's not already a SQL database, create a new SQL database and use the `dbWriteTable()` function to add at minimum three tables to the SQL database.**

    ```{r}
    flight_data  <- dbConnect(RSQLite::SQLite(), "nycflights13.db")

    # Create SQLite database
    dbWriteTable(flight_data, "flights", flights, overwrite = TRUE)
    dbWriteTable(flight_data, "weather", weather, overwrite = TRUE)
    dbWriteTable(flight_data, "airports", airports, overwrite = TRUE)

    dbListTables(flight_data)

    # View dataset variables 
    flights_columns <- dbListFields(flight_data, "flights")
    weather_columns <- dbListFields(flight_data, "weather")
    airports_columns <- dbListFields(flight_data, "airports")

    # Display the column names
    print(flights_columns)
    print(weather_columns)
    print(airports_columns)
    ```

::: callout-important
## Important

The output has been limited to display only the first 10 rows to improve readability. If you'd like to view the entire dataset, please remove `n = 10` from the end of each code block.
:::

2.  **Write three SQL queries using `dbGetQuery()` to explore and summarize the data. You must use at least five different SQL functions (e.g. `SELECT`, `GROUP BY`, `DISTINCT`, `SUM`, etc).**

    ```{r}
    # QUERY 1: The number of flights by airline and delays in NY in 2013 by season  
    query1 <- "
      SELECT 
        carrier AS airline,
        CASE
          WHEN month IN (12, 1, 2) THEN '1' -- Winter
          WHEN month IN (3, 4, 5) THEN '2' -- Spring 
          WHEN month IN (6, 7, 8) THEN '3' -- Summer
          WHEN month IN (9, 10, 11) THEN '4' -- Fall
        END AS season,
        COUNT(*) AS total_flights,
        SUM(CASE WHEN dep_delay > 0 THEN 1 ELSE 0 END) AS delayed_flights
      FROM flights
      GROUP BY carrier, season
    "

    query_result1 <- dbGetQuery(flight_data, query1)
    head(query_result1, n=10)
    ```

    ```{r}
    # QUERY 2: The average number of departure delay minutes by airlines in NY in 2013
    query2 <- "
    SELECT 
        carrier AS airline,
        AVG(dep_delay) AS avg_delay_minutes
    FROM flights
    WHERE dep_delay IS NOT NULL
    GROUP BY carrier
    ORDER BY avg_delay_minutes DESC
    "

    query_result2 <- dbGetQuery(flight_data, query2)
    head(query_result2, n=10)
    ```

    ```{r}
    # QUERY 3: Departure delay minutes by airline across seasons
    query3 <- "
    SELECT 
        carrier AS airline,
        CASE
          WHEN month IN (12, 1, 2) THEN '1' -- Winter
          WHEN month IN (3, 4, 5) THEN '2' -- Spring 
          WHEN month IN (6, 7, 8) THEN '3' -- Summer
          WHEN month IN (9, 10, 11) THEN '4' -- Fall
        END AS season,
        AVG(dep_delay) AS avg_delay_minutes
    FROM flights
    WHERE dep_delay IS NOT NULL
    GROUP BY carrier, season
    ORDER BY season, avg_delay_minutes DESC
    "

    query_result3 <- dbGetQuery(flight_data, query3)
    head(query_result3, n=10)
    ```

3.  **Write two SQL queries to create new features from existing data. For example, if using `nycflights13`, you could think about how to**

    -   **Bin departure times into time-of-day categories (e.g., morning, afternoon, evening).**

    -   **Lag features like the previous day's average delay by carrier, which can be helpful for predictions.**

    -   **Merge additional weather data (such as hourly temperature, precipitation, etc., if available). e.g. this could be done using SQL joins.**

        ```{r}
        # QUERY 1: Binning departure times into time-of-day categories
        query_time_of_day <- "
          SELECT *,
                 CASE
                   WHEN dep_time BETWEEN 0 AND 600 THEN 'Early Morning'
                   WHEN dep_time BETWEEN 601 AND 1200 THEN 'Morning'
                   WHEN dep_time BETWEEN 1201 AND 1700 THEN 'Afternoon'
                   WHEN dep_time BETWEEN 1701 AND 2400 THEN 'Evening'
                   ELSE 'Unknown'
                 END AS time_of_day
          FROM flights
        "
        flights_time_of_day <- dbGetQuery(flight_data, query_time_of_day)

        flights_time_of_day <- flights_time_of_day %>% 
          filter(complete.cases(.)) %>%
          select(dep_delay, carrier, time_of_day)

        head(flights_time_of_day, n=10)
        ```

        ```{r}
        # QUERY 2: Creating a delay category based on departure delay
        query_delay_category <- "
          SELECT *,
                 CASE
                   WHEN dep_delay <= 0 THEN 'No Delay'
                   WHEN dep_delay > 0 AND dep_delay <= 30 THEN 'Minor Delay'
                   WHEN dep_delay > 30 THEN 'Major Delay'
                   ELSE 'Unknown'
                 END AS delay_category
          FROM flights
        "

        flights_delay_category <- dbGetQuery(flight_data, query_delay_category)

        flights_delay_category <- flights_delay_category %>%
          filter(complete.cases(.)) %>%
          select(dep_delay, carrier, delay_category)

        head(flights_delay_category, n=10)
        ```

4.  **Visualize your data by creating a plot with `ggplot2`. For example, if using `nycflights13`, you could think about how to visualize delays by month, carrier, or weather conditions.**

    ```{r}
    # Results 1: The number of flights by airline and season
    query_result1$season <- factor(
      query_result1$season, 
      levels = c("1", "2", "3", "4"), 
      labels = c("Winter", "Spring", "Summer", "Fall")
    )

    ggplot(query_result1, aes(x = airline, y = total_flights, fill = season)) +
      geom_bar(stat = "identity", position = "dodge") +
      labs(
        title = "Number of Flights by Airline and Season",
        subtitle = "Data based on New York in 2013",
        caption = "Graphed by Amanda Kim & Da Yea Song",
        x = "Airline",
        y = "Number of Flights",
        fill = "Season"
      ) +
      scale_fill_viridis_d(option = "D") +
      theme_minimal() +
      theme(
        plot.title = element_text(face = "bold", hjust = 0.5),
        plot.subtitle = element_text(face = "italic", hjust = 0.5),
        axis.title.x = element_text(face = "bold"),
        axis.title.y = element_text(face = "bold")
      )
    ```

    ```{r}
    # Results 2: Create a plot for average departure delay by airline
    ggplot(query_result2, aes(x = airline, y = avg_delay_minutes)) +
      geom_bar(stat = "identity", position = "dodge", fill = "darkslategray3") +
      labs(
        title = "Average Departure Delay by Airline",
        subtitle = "Data based on New York in 2013",
        caption = "Graphed by Amanda Kim & Da Yea Song",
        x = "Airline",
        y = "Average Departure Delay (minutes)"
      ) +
      theme_minimal() +
      theme(
        plot.title = element_text(face = "bold", hjust = 0.5),
        plot.subtitle = element_text(face = "italic", hjust = 0.5),
        axis.title.x = element_text(face = "bold"),
        axis.title.y = element_text(face = "bold")
      )
    ```

    ```{r}
    # Results 3: Average delays for each airline across seasons
    query_result3$season <- factor(
      query_result3$season, 
      levels = c("1", "2", "3", "4"), 
      labels = c("Winter", "Spring", "Summer", "Fall")
    )

    ggplot(query_result3, aes(x = airline, y = avg_delay_minutes, fill = season)) +
      geom_bar(stat = "identity", position = "dodge", alpha = 0.7) +    labs(
        title = "Average Departure Delay by Airline Across Seasons",
        subtitle = "Data based on New York in 2013",
        caption = "Graphed by Amanda Kim & Da Yea Song",
        x = "Airline",
        y = "Average Departure Delay (Minutes)",
        fill = "Season"
      ) +
      theme_minimal() +
      theme(
        plot.title = element_text(face = "bold", hjust = 0.5),
        plot.subtitle = element_text(face = "italic", hjust = 0.5),
        axis.title.x = element_text(face = "bold"),
        axis.title.y = element_text(face = "bold"),
        legend.position = "bottom"
      )
    ```

    ```{r}
    # Query 1: Time-of-Day Categories
    flights_time_of_day$time_of_day <- factor(
      flights_time_of_day$time_of_day,
      levels = c("Early Morning", "Morning", "Afternoon", "Evening")
    )

    ggplot(flights_time_of_day, aes(x = time_of_day)) +
      geom_bar(fill = "lightsalmon") +
      labs(
        title = "Number of Flights by Time of Day",
        subtitle = "Data based on New York in 2013",
        caption = "Graphed by Amanda Kim and Da Yea Song",
        x = "Time of Day",
        y = "Number of Flights"
      ) +
      theme_minimal() + 
      theme(
        plot.title = element_text(face = "bold", hjust = 0.5),
        plot.subtitle = element_text(face = "italic", hjust = 0.5),
        axis.title.x = element_text(face = "bold"),
        axis.title.y = element_text(face = "bold")
      )
    ```

    ```{r}
    # Query 2: Flight Delay Categories
    ggplot(flights_delay_category, aes(x = delay_category)) +
      geom_bar(fill = "mediumaquamarine") +
      labs(
        title = "Number of Flights by Delay Category",
        subtitle = "Data based on New York in 2013",
        caption = "Graphed by Amanda Kim and Da Yea Song",
        x = "Delay Category",
        y = "Number of Flights"
      ) +
      theme_minimal() + 
      theme(
        plot.title = element_text(face = "bold", hjust = 0.5),
        plot.subtitle = element_text(face = "italic", hjust = 0.5),
        axis.title.x = element_text(face = "bold"),
        axis.title.y = element_text(face = "bold")
      )
    ```

5.  **Report your findings. Provide a paragraph summarizing your methods and key findings. Include any limitations or potential biases in your analysis. Be sure to comment and organize your code so is easy to understand what you are doing.**

    Across all airlines, it seems like season #3 ("Summer") is, on average, the highest in departure delay (in minutes) compared to other seasons. Comparing all 16 airlines, F9, EV, YV have the highest average departure delays. US, HA, and AS airlines have the lowest departure delays (in minutes). When comparing all airlines across seasons, OO airline had the largest average departure delay during the Winter and Spring (close to 60-70 minutes). It's important to highlight that HA airlines may have had relatively high average departure delays during the winter, in the Spring and Fall, HA airlines have had a negative average departure delay, signifying that the airline is departing early. 

    [**Limitations in Analysis:**]{.underline} 

-   **Dataset Scope**: The dataset only includes flights departing from NYC airports, which means that our analysis is not representative of airlines' performance nationwide or globally.

-   **Sample Size**: Airlines with fewer flights in the dataset might have averages that are not as reliable or statistically significant compared to airlines with a larger number of flights.

-   **External Factors to Consider**: Delays may not be entirely within the airline's control. Factors like airport congestion, air traffic control delays, or weather at origin/destination airports can all contribute to departure delays.

    [**Potential Biases:**]{.underline}

-   **Airport-Specific Bias:** Since the dataset focuses only on NYC airports, delays may be influenced more by NYC airport operations, infrastructure, or congestion, rather than the airlines themselves.

-   **Weather Bias**: NYC airports are heavily affected by winter weather, potentially biasing results for airlines with many winter flights departing from NYC. 

-   **Excluded Flights**: Flights canceled or diverted due to extreme weather or other factors are excluded from delay calculations, potentially underestimating the operational challenges faced by certain airlines or in certain seasons.

***Data cleaning*****:** <br> - `dbWriteTable()` used to write data from R into a table in a database <br> - `SELECT()` specifies which columns of data we want to retrieve from a database table <br> - `GROUP BY` used to group rows that have the same values in specified columns into summary rows <br> - `ORDER BY` sorts the result set by one or more columns in ascending (`ASC`) or descending (`DESC`) order. <br> - `SUM` calculates the total (sum) of a numeric column <br> - `CASE` implements conditional logic within a query, similar to an `IF` or `SWITCH` statement in programming <br> - `COUNT` counts the number of rows that match a condition, or counts non-NULL values in a column <br>

***Visualization*****:** <br> - `geom_bar()` to plot the 5 plots looking at average minutes delayed across seasons and by airline

## Part 2

In this part, you and your partner will use use `caret` or `tidymodels` to train a predictive model for a task or outcome, perform model evaluation, and summarize your findings. For this part, you can use the data directly in R as standard dataframes or tibbles. You do not need to build the models In this section, you must do the following when building your prediction model:

**Research Question:** Can weather conditions predict departure delays for American Airlines (AA) from NYC in 2013?

**Data Set Up**

```{r}
data(flights)
data(weather)
```

```{r}
# Select flight details for American Airlines (AA)
aa_flights <- flights %>%
  filter(carrier == "AA") %>%
  select(dep_delay, distance, month, day)

# Select weather conditions from the dataset
weather <- weather %>%
  select(month, day, visib, precip, wind_speed)

# Calculate the average of weather conditions per day 
weather_aggregated <- weather %>%
  group_by(month, day) %>%
  summarize(
    visib = mean(visib, na.rm=TRUE), 
    precip = mean(precip, na.rm=TRUE),
    wind_speed = mean(wind_speed, na.rm=TRUE),
    .groups = "drop"
  )
```

::: callout-note
## Note

The average daily weather conditions were calculated since weather data was recorded at multiple time points throughout each day.
:::

```{r}
# Merge flight details with weather conditions 
aa_data <- aa_flights %>%
  left_join(weather_aggregated, by = c("month", "day"))

# Create two new variables (delay_status and season)
aa_data <- aa_data %>%
  mutate(
    delay_status = ifelse(dep_delay <= 0, 0, 1), # 0 (no delay), 1 (delay)
    season = case_when(
      month %in% c(12, 1, 2) ~ 1, # Winter 
      month %in% c(2, 3, 4) ~ 2, # Spring 
      month %in% c(5, 6, 7) ~ 3, # Summer
      month %in% c(8, 9, 10) ~ 4, # Fall 
    )
  )

# Remove any rows with missing values 
aa_data <- na.omit(aa_data)
```

1.  **Split the data into training and test sets.**

    ```{r}
    set.seed(100)

    train <- createDataPartition(aa_data$delay_status, p = 0.8, list = FALSE)

    train_data <- aa_data[train, ]
    test_data <- aa_data[-train, ]

    dim(train_data)  
    dim(test_data) 
    ```

2.  **Choose a machine learning model appropriate for the task (e.g. consider if it's a binary or continuous outcome and choose an appropriate model). For example, if you are using the `nycflights13` dataset, you could could pick**

    -   **Logistic Regression (if predicting delay as a binary outcome, e.g., delay/no delay).**

    -   **Linear Regression (if predicting the length of delay in minutes).**

    -   **Decision Trees or Random Forests (for both binary classification or regression).**

        Since the outcome involves predicting whether an AA flight will be delayed or not, we are working with a binary classification task. Therefore, we selected a logistic regression model.

        ```{r}
        train_data$delay_status <- as.factor(train_data$delay_status)
        test_data$delay_status <- as.factor(test_data$delay_status)
        ```

3.  **Train the model using `caret` or `tidymodels` in R using the training data.**

    ```{r}
    # Train a logistic regression model 
    model <- train(
      delay_status ~ distance + visib + precip + wind_speed + season, 
      data = train_data, 
      method = "glm",
      family = "binomial",
      trControl = trainControl(method = "cv", number = 5)
    )

    summary(model)
    model$results
    ```

4.  **Assess the model performance both the training and test datasets using metrics like Accuracy or area under the curve (AUC) (if classification) or root mean squared error (RMSE) (if regression).**

    ```{r}
    # Evaluate the training dataset 
    train_evaluate <- predict(model, newdata = train_data)

    train_confusion_matrix <- confusionMatrix(train_evaluate, train_data$delay_status)
    print(train_confusion_matrix)

    # Calculate AUC for the training dataset
    train_prob <- predict(model, newdata = train_data, type = "prob")[, 2]

    train_roc <- roc(train_data$delay_status, train_prob)
    train_auc <- auc(train_roc)
    cat("Training AUC:", train_auc, "\n")
    ```

    ```{r}
    # Evaluate the testing dataset 
    test_evaluate <- predict(model, newdata = test_data)

    test_confusion_matrix <- confusionMatrix(test_evaluate, test_data$delay_status)
    print(test_confusion_matrix)

    # Calculate AUC for the testing dataset
    test_prob <- predict(model, newdata = test_data, type = "prob")[, 2]

    test_roc <- roc(test_data$delay_status, test_prob)
    test_auc <- auc(test_roc)
    cat("Testing AUC:", test_auc, "\n")
    ```

5.  **Finally, report your findings. For example, if you are predicting flight delays, provide 1-2 practical recommendations for reducing delays. Broadly, provide a paragraph summarizing your methods and key findings. Include limitations or potential biases in training and evaluation your machine learning model. Be sure to comment and organize your code so is easy to understand what you are doing.**

    NYC flight data from 2013 was used to predict the flight delay status for American Airlines (AA). Various weather conditions, along with flight distance and season, were selected as predictor variables for delay status. The data was split into training and testing datasets, and the model was trained using cross-validation. Given the binary nature of the outcome (delay vs. no delay), a logistic regression model was employed.

    Using the average measurements of visibility, precipitation, wind speed, flight distance, and season, the model achieved an accuracy of 67.8% on the training dataset, with a sensitivity of 98.3% and a specificity of 3.6%. On the testing dataset, the model achieved an accuracy of 67.2%, a sensitivity of 98.3%, and a specificity of 4.1%. The area under the curve (AUC) for the training and testing datasets was 59.1 and 59.0, respectively, suggesting that the model's ability to predict the delay status of AA flights is only slightly better than random guessing.

    [Recommendations Based on Findings:]{.underline}

    **Weather Preparedness**: During winter months, AA should improve de-icing operations and staff scheduling to minimize weather-related delays and staff shortages/call outs. Airlines/airports should also look into more advanced weather-monitoring technologies to optimize flight schedules during adverse conditions.

    [Limitations or Potential biases:]{.underline}

<!-- -->

1.  **Sample Bias**: The model relies on the **2013 NYC flight dataset**, which does not reflect current patterns or conditions, especially with changes in airline operations, technology, and infrastructure during the last decade.

2.  **Other Variables to Add/Consider**: The data did not include additional factors, such as staffing levels and air traffic control delays, which may limit the model's predictive accuracy. To improve the model's AUC/predictive accuracy, we think it's important that we consider and add additional variables into the model.

3.  **Class Imbalance**: If delayed flights are underrepresented in the data, the model might not be able to accurately predict delays, which would ultimately skew results toward the majority class (on-time flights). Looking at part 1, we can see that the average departure delay for AA is under 10 minutes. If the number of delayed flights are "minimal" compared to "on time" flights, than the results might skew towards "on-time" flights, ultimately, affecting our model's ability to better predict departure delays.

***Data cleaning*****:** <br> - `filter()` to select data from AA <br> - `select()` to select specific variables of interest <br> - `group_by()` to group weather data by month and day <br> - `summarize` to calculate the mean measurements for the weather variables <br> - `left_join()` to merge flight and weather datasets <br> - `mutate()` to create new variables <br> <br>

***Data training*****:** <br> - `createDataPartition` to split the data into training and test sets <br> -`train()` to train the model<br> -`predict()` to evaluate the model<br> -`roc` to calculate the area under the curve
